---
title: "Insurance Claim Prediction"
editor: visual
code-fold: true
categories: [Random Forest, Bagging, Boosting, Regression Tree]
#engine: knitr
---

```{r setup, echo=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  comment = ""
)
library(kableExtra)
library(caret)
library(caTools)
library(doParallel) # for parallel backend to foreach
# library(foreach)
# # for parallel processing with for loops
library(randomForest)
library(rpart)
library(rpart.plot)
library(gbm)

# Setup parallel backend
num_cores <- parallel::detectCores()- 1
cl <- makeCluster(num_cores)
registerDoParallel(cl)

options(ggplot2.continuous.colour = "viridis",
ggplot2.continuous.fill = "viridis")
theme_set(theme_minimal())
theme_update(panel.grid.minor = element_blank())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(corrplot::corrplot)
set.seed(1)
```

```{css, echo = FALSE}
.justify {
  text-align: justify !important
}
```


```{r}
load("E:/Personal Website/blogs/claim prediction/telematics.RData")
dat_1 <- dat_1 |>
  mutate(across(where(is.character), as.factor))
dat_1$AMT_Claim <- scale(dat_1$AMT_Claim)

```

## Project Goal

The main goal is how to develop a model that predicts the amount of claim(`AMT Clain`) using `telematics.RData`. For additional information regarding the data, refer to [https://www.mdpi.com/2227-9091/9/4/58](https://www.mdpi.com/2227-9091/9/4/58).


## Exploratory Data Analysis 

There are no missing values. We found 84 observations with negative car age. Negative car age does not make sense so we exclude these observations. Using the `findCorrelation()` function in R, we identify variables highly correlated($>= 0.8$) with other variables. The following variables are then excluded. We also exclude the number of claims during observations(`NB_claim`). `NB_claim` is considered as a response in the original article.

```{r}
tele_df <- dat_1 |> dplyr::filter(Car.age >= 0) |>
  mutate(across(where(is.character), as.factor))

#percentage of missing values
naniar::pct_miss(tele_df)

num_variables <- select(tele_df, where(is.numeric)) |> select(-AMT_Claim)
cor_matrix <- cor(num_variables, use = "complete.obs")
high_cor_vars <- findCorrelation(cor_matrix, cutoff = .8, names = TRUE)

matrix(high_cor_vars,  nrow = 4, ncol = 5, byrow = TRUE) |> kableExtra::kable()
```


```{r}
tele_df_clean <- select(tele_df, -any_of(c(high_cor_vars, "NB_Claim")))
```


My final dataset for modeling has 32 variables with 3780 observations.

```{r}
summarise(tele_df_clean,
          across(
            where(is.numeric),
            .fns = list(
              min = min,
              median = median,
              mean = mean,
              stdev = sd,
              q25 = ~ quantile(., 0.25),
              q75 = ~ quantile(., 0.75),
              max = max
            ),
            .names = "{.col}_{.fn}"
          )) |>
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "stat"),
    names_pattern = "^(.*)_(min|median|mean|stdev|q25|q75|max)$"
  ) |>
  pivot_wider(names_from = stat, values_from = value) |>
  arrange(variable) |> #alphabetical order
  kable(digits = 3, caption = "Summary statistics") |>
  kable_styling(full_width = TRUE,
                bootstrap_options = c("striped", "hover"))
```


```{r}
tele_df_clean |>
  select(where(is.character), where(is.factor)) |>
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "level") |>
  group_by(variable, level) |>
  summarise(count = n(), .groups = "drop_last") |>
  mutate(prop = count / sum(count)) |> ungroup() |>
  arrange(variable, desc(count)) |>
  kable(digits = 3, caption = "Frequency and Proportion of Categorical Variables") |>
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover"))
```


```{r}
#| fig-cap: "Histogram of aggregated amout claimed"
ggplot(tele_df_clean, aes(x = AMT_Claim)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black",
    alpha = 0.7
  )
```

The distribution of standardized claims is right skewed.

```{r}
split <- sample.split(tele_df_clean$AMT_Claim, SplitRatio = .8)
train_df <- subset(tele_df_clean,split == TRUE)
test_df <- subset(tele_df_clean, split == FALSE)
cat("Train data sample size:", nrow(train_df))
cat("Test data sample size:", nrow(test_df))

```

## Models 

### Regression Tree
```{r}
insur_tree <- rpart(AMT_Claim ~ .,
                    data = train_df,
                    method = "anova",
                    "cp" = .008)
rpart.plot(insur_tree)

insur_tree_importance <- with(insur_tree, tibble(
  variable = names(variable.importance),
  importance = as.numeric(variable.importance)
)) |> arrange(desc(importance)) |> slice_head(n = 100)

ggplot(insur_tree_importance, 
       aes(x = factor(variable, levels = rev(variable)), y = importance)) +
  geom_col(fill = "steelblue") +
  labs(x = "variable") +
  coord_flip()

amt_claim_hat <- predict(insur_tree, newdata = test_df)
insur_treemse <- mean((test_df$AMT_Claim - amt_claim_hat)^2)
cat("MSE of the regression tree before pruning:",
    insur_treemse |> round(3))
cptable <- insur_tree$cptable
ggplot(cptable, aes(nsplit, xerror)) +
  geom_point() +
  geom_line()


```

```{r}
#prune the trees
ins_bestcp <- cptable[which.min(cptable[, "xerror"]), "CP"]
insur_pruntree <- prune(insur_tree, cp = ins_bestcp)

rpart.plot(insur_pruntree)

with(insur_pruntree,
     tibble(
       variable = names(variable.importance),
       importance = as.numeric(variable.importance)
     )) |>
  ggplot(aes(x = factor(variable, levels = rev(variable)), y = importance)) +
  geom_col(fill = "steelblue") +
  labs(x = "variable") +
  coord_flip()

#after prunning prediction
prune_amtclaim_hat <- predict(insur_pruntree, newdata = test_df)
prune_mse <- mean((test_df$AMT_Claim - prune_amtclaim_hat)^2)
cat("MSE after pruning:", prune_mse)
mse_dif <- ((insur_treemse - prune_mse) / (insur_treemse)) * 100
cat("MSE reduction:", mse_dif |> round(3))
```



### Bagging
```{r}
#---------------------choosing best ntrees----------------
#| fig-cap: "Optimal number of trees for bagging"
#---------------------trees to grow--------------------------

ntree_values <- c(50, 100, 200, 300, 500)
# # Run cross-validation in parallel
# bagging_cverrors <- foreach(n = ntree_values,
#                             .combine = rbind,
#                             .packages = "caret") %dopar% {
#                               model <- train(
#                                 AMT_Claim ~ .,
#                                 data = train_df,
#                                 method = "rf",
#                                 trControl = trainControl(method = "cv", number = 10),
#                                 ntree = n,
#                                 importance = TRUE
#                               )
# 
# data.frame(ntree = n, RMSE = min(model$results$RMSE))
# }
# bagging_cverrors <- write_csv(bagging_cverrors, "bagging_cverrors")
# bagging_cverrors <- read_csv("bagging_cverrors")
# ggplot(bagging_cverrors, aes(ntree, RMSE))+
# geom_point()+
# geom_line()
# #
# # bagging_bestntree <- bagging_cverrors$ntree[which.min(bagging_cverrors$RMSE)]

```

```{r}
# #-----------------bagging model---------------------------------
# #---------------------------------------------------------------------
insur_bagging <- randomForest(
  AMT_Claim ~ .,
  data = train_df,
  importance = TRUE,
  ntree = 500,
  mtry = floor(ncol(train_df) - 1)
)
# varImpPlot(insur_bagging, main = "")

#variable importance plot using ggplot
importance(insur_bagging) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename(importance= `%IncMSE`) %>%
  arrange(desc(importance)) %>%
  ggplot(aes(x = reorder(variable, importance), y = importance))+
  geom_bar(stat = "identity", fill = "steelblue")+
  coord_flip()

insurbag_hat <- predict(insur_bagging, newdata = test_df)
bagging_mse <- mean((test_df$AMT_Claim- insurbag_hat)^2)
cat("Bagging MSE:", bagging_mse)
```


### Boosting
```{r}
#---------------------------boosting ntree search--------------------------------------------
# boosting_cv <- foreach(n = ntree_values,
#                        .combine = rbind,
#                        .packages = "gbm") %dopar% {
#                          set.seed(123)
#                          model <- gbm(
#                            AMT_Claim ~ .,
#                            data = train_df,
#                            distribution = "gaussian",
#                            n.trees = n,
#                            shrinkage = 0.01,
#                            interaction.depth = 1,
#                            n.minobsinnode = 10,
#                            cv.folds = 10,
#                            verbose = FALSE
#                          )
#                          
#                          data.frame(n_tree = n, cv_error = min(model$cv.error))
#                        }
# 
# boosting_cv |>
# ggplot(aes(n_tree, cv_error))+
# geom_point()+
# geom_line()
# boosting_bestntree <- boosting_cv$n_tree[which.min(boosting_cv$cv_error)]
insur_boosting <- gbm(
  AMT_Claim ~ .,
  data = train_df,
  distribution = "gaussian",
  n.trees = 500,
  interaction.depth = 1
)
insur_boostingsummary <- summary(insur_boosting, plotit = FALSE)
ggplot(insur_boostingsummary, aes(x = reorder(var, rel.inf), y = rel.inf)) +
  geom_col(fill = "steelblue") +
  coord_flip()+
  labs(x = "Variable", y = "Relative Influence")
#---------------------------boosting prediction--------------------------
boost_amthat <- predict(insur_boosting, newdata = test_df)
boosting_insurmse <- mean((test_df$AMT_Claim - boost_amthat)^2)
cat("Boosting MSE:", boosting_insurmse)


```

### Random Forest
```{r}
rf_claim <- randomForest(
  AMT_Claim ~ .,
  data = train_df,
  importance = TRUE,
  mtry = floor(sqrt(ncol(train_df) - 1)),
  ntree = 500
)
# 
# varImpPlot(rf_claim, main = "")

#variable importance with ggplot
importance(rf_claim) %>% as.data.frame() %>%
  mutate(variable = rownames(.)) %>%
  rename(importance= `%IncMSE`) %>%
  arrange(desc(importance)) %>%
  ggplot(aes(x = reorder(variable, importance), y = importance))+
  geom_bar(stat = "identity", fill = "steelblue")+
  coord_flip()

#predict on test set and compute mse
rf_yhat <- predict(rf_claim, newdata = test_df)
rf_mse <- mean((test_df$AMT_Claim - rf_yhat)^2)
cat("RandomForest MSE:", rf_mse)

```


## Model Comparison
```{r}
tibble(
  method = c("Pruned regression tree", "Bagging", "Boosting", "RandomForest"),
  mse = c(prune_mse, bagging_mse, boosting_insurmse, rf_mse)
) |>
  kable(digits = 3, caption = "Model comparison")
```
